{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-07-07 thesis/education.tex
describe mutex bug

\begin{table}[h]
	\begin{center}
		% TODO: calc p values for this
		% note, smth like fisher's method which "combines independent studies into a meta p value"
		% cannot be applied for 2 reasons - 1, these are not independent since the same studence were
		% running landslide on diff test cases; 2, the hypotheses tested are different for each test
	\begin{tabular}{r||cc|cc|c}
		& \multicolumn{2}{c|}{Landslide users} & \multicolumn{2}{c|}{Non-users} & \\
		Bug name & \#correct & total & \#correct & total & $\Delta P(\mathsf{correct})$ \\
		\hline
		%Exit UAF  & 9	& 21	& 2	& 16	& $+30.4\%$ \\ % not filtering out did not run tej; also using "old" thing..?
		Exit UAF  & 11	& 21	& 0	& 16	& $+52.4\%$ \\
		% nb. this is not filter out the did not run paras -- smae for all of the above/below
		% Paraguay  & 22	& 24	& 12	& 13	& $-0.64\%$ \\ % not filtering out did not run para
		Paraguay  & 20	& 21	& 14	& 16	& $+07.7\%$ \\
		% nb this is any downgrade related ding not just dgr tested ones
		% nb 1 is msising bc didn't impl rwlock at all
		%Downgrade & 17	& 24	& 8	& 12	& $+4.17\%$ \\ % not filtering out either did not run, or unrel ding
		Downgrade & 10	& 13	& 19	& 23	& $-05.7\%$ \\ % TODO verify these buggers by eye
		% I spent more hours than I should have looking at this landslide-user-submitted-bug group.
		% Can't name the group here, but if you're me and have the analysis file,
		% it's the one from f17 with 3 mutex_test logs in their "u" snapshot.
		% It seems like a mysterious older version of mutex.c was present between their "a" version
		% (which gets fully verified very quickly btw) and their "u" version (for which landslide finds
		% the bug in just a few minutes, no way they could be running the same version; plus the reported
		% data races and so on in the id-log don't line up to the "u" mutex.c source code).
		% That version was correct, then they updated it (introducing the bug) and ran a bunch of other tests,
		% but didn't rerun mutex_test to check their update.
		Mutex     & 17	& 18	& 19	& 19	& $-05.6\%$ \\
		\hline
		% TODO: figure out if calculating a p value (or w/e) for this the same way is valid,
		% since they summands are arguably not independent? ..or maybe they are?
		Total     & 58	& 73	& 52	& 74	& $+09.2\%$ \\
	\end{tabular}
	\end{center}
	\caption{
		Correlation of student Landslide use with solving certain bugs in their final submission during Fall 2017 and Spring 2018 semesters.
		Students who used Landslide in general but whose usage snapshots showed they did not run the test case in question are counted as non-Landslide-users here.
		$\Delta P$ stands for the increased probability that a student submits a correct P2
		(with respect to each bug)
		given that they use Landslide, versus not using Landslide; i.e.,
		$P(\mathsf{correct}|\mathsf{landslide})-
		P(\mathsf{correct}|\mathsf{non\text{-}user})$.
		Note that the bottom row's totals double-count students,
		which only makes sense if you believe the incidences of each bug within a given submission are independent.
		}
	\label{tab:eval-common-bugs}
\end{table}

{\bf Thread-exit use-after-free.}
%During {\tt thr\_exit()}, a thread must arrange for its own stack memory allocation to be freed.
After a thread finishes exiting, the memory allocated for its stack must be reclaimed for other uses.
A common student pitfall is to allow a {\tt thr\_join()}ing thread to free said stack space
while {\tt thr\_exit()} is still executing userspace C code (which typically accesses the stack),
or even for {\tt thr\_exit()} to free it itself.
In the former case, threads must interleave specifically to exhibit a use-after-free;
in the latter case, the use-after-free will be deterministic (i.e., present in all interleavings).
Other than Landslide, however,
the students have no Valgrind-like heap debugging tool which would report a bug immediately upon any illegal heap access.
This means that a subsequent {\tt thr\_create()} invocation would need to race to recycle the memory for a new thread stack
and conflict with the old exiting thread
before any problem could be detected.
% TODO: mark which of the tej bugs were deterministic vs race
The Landslide-friendly test {\tt thr\_exit\_join} is most likely to expose this bug.
%, and
%the official rubric penalties for it are
%{\tt :exit\_stack\_minor -2} and
%{\tt :exit\_stack\_race -4} (for the former case),
%and
%{\tt :uses\_freed\_stack -4} (for the latter).
Whether each student's submitted P2 solved or suffered from this bug
was determined by me personally inspecting their code.

@@

{\bf R/W-lock downgrade.}
In addition to the standard R/W-lock interface,
P2 requires students to implement {\tt rwlock\_downgrade()}:
called with the lock held in write mode,
the caller adopts the reader role instead,
allowing other waiting reader threads to proceed simultaneously,
all while allowing no waiting writers to access in between.
%
The Landslide-friendly test {\tt rwlock\_downgrade\_read\_test}
checks that readers are allowed simultaneous access after a downgrade with no possibility for deadlock.
Rather than one specific bug, this refers to any of several failures that can arise during a downgrade.
%, and
%the official rubric penalties related to this bug are
%{\tt :downgrade\_no\_signal -1},
%%{\tt :unsafe\_rwlock\_downgrade -2}, % don't think this is right; not relevant anyway (overlaps)
%{\tt :downgrade\_partial\_signal -1}, % not sure if this means smth that dgr would expose
%{\tt :downgrade\_\allowbreak{}lies -3}, and
%{\tt :downgrade\_deadlock -4}.
Whether each student's submitted P2 solved or suffered from these bugs
was determined by checking their grade files for any of the above-listed penalties
(which were assessed by the TAs as a result of their manual inspection, rather than mine).

>>>>

\begin{table}[h]
	\begin{center}
		% TODO: calc p values for this
		% note, smth like fisher's method which "combines independent studies into a meta p value"
		% cannot be applied for 2 reasons - 1, these are not independent since the same studence were
		% running landslide on diff test cases; 2, the hypotheses tested are different for each test
	\begin{tabular}{r||cc|cc|c}
		& \multicolumn{2}{c|}{Landslide users} & \multicolumn{2}{c|}{Non-users} & \\
		Bug name & \#correct & total & \#correct & total & $\Delta P(\mathsf{correct})$ \\
		\hline
		%Exit UAF  & 9	& 21	& 2	& 16	& $+30.4\%$ \\ % not filtering out did not run tej; also using "old" thing..?
		Exit UAF  & 11	& 21	& 0	& 16	& $+52.4\%$ \\
		% I spent more hours than I should have looking at this landslide-user-submitted-bug group.
		% Can't name the group here, but if you're me and have the analysis file,
		% it's the one from f17 with 3 mutex_test logs in their "u" snapshot.
		% It seems like a mysterious older version of mutex.c was present between their "a" version
		% (which gets fully verified very quickly btw) and their "u" version (for which landslide finds
		% the bug in just a few minutes, no way they could be running the same version; plus the reported
		% data races and so on in the id-log don't line up to the "u" mutex.c source code).
		% That version was correct, then they updated it (introducing the bug) and ran a bunch of other tests,
		% but didn't rerun mutex_test to check their update.
		Mutex     & 17	& 18	& 19	& 19	& $-05.6\%$ \\
		% nb. this is not filter out the did not run paras -- smae for all of the above/below
		% Paraguay  & 22	& 24	& 12	& 13	& $-0.64\%$ \\ % not filtering out did not run para
		Paraguay  & 20	& 21	& 14	& 16	& $+07.7\%$ \\
		% nb this is any downgrade related ding not just dgr tested ones
		% nb 1 is msising bc didn't impl rwlock at all
		%Downgrade & 17	& 24	& 8	& 12	& $+4.17\%$ \\ % not filtering out either did not run, or unrel ding
		Downgrade & 10	& 13	& 19	& 23	& $-05.7\%$ \\ % TODO verify these buggers by eye
		\hline
		% TODO: figure out if calculating a p value (or w/e) for this the same way is valid,
		% since they summands are arguably not independent? ..or maybe they are?
		Total     & 58	& 73	& 52	& 74	& $+09.2\%$ \\
	\end{tabular}
	\end{center}
	\caption{
		Correlation of student Landslide use with solving certain bugs in their final submission during Fall 2017 and Spring 2018 semesters.
		Students who used Landslide in general but whose usage snapshots showed they did not run the test case in question are counted as non-Landslide-users here.
		$\Delta P$ stands for the increased probability that a student submits a correct P2
		(with respect to each bug)
		given that they use Landslide, versus not using Landslide; i.e.,
		$P(\mathsf{correct}|\mathsf{landslide})-
		P(\mathsf{correct}|\mathsf{non\text{-}user})$.
		Note that the bottom row's totals double-count students,
		which only makes sense if you believe the incidences of each bug within a given submission are independent.
		}
	\label{tab:eval-common-bugs}
\end{table}

{\bf Thread exit/join.}
%During {\tt thr\_exit()}, a thread must arrange for its own stack memory allocation to be freed.
After a thread finishes exiting, the memory allocated for its stack must be reclaimed for other uses.
A common student pitfall is to allow a {\tt thr\_join()}ing thread to free said stack space
while {\tt thr\_exit()} is still executing userspace C code (which typically accesses the stack),
or even for {\tt thr\_exit()} to free it itself.
In the former case, threads must interleave specifically to exhibit a use-after-free;
in the latter case, the use-after-free will be deterministic (i.e., present in all interleavings).
Other than Landslide, however,
the students have no Valgrind-like heap debugging tool which would report a bug immediately upon any illegal heap access.
This means that a subsequent {\tt thr\_create()} invocation would need to race to recycle the memory for a new thread stack
and conflict with the old exiting thread
before any problem could be detected.
% TODO: mark which of the tej bugs were deterministic vs race
The Landslide-friendly test {\tt thr\_exit\_join} is most likely to expose this bug.
%, and
%the official rubric penalties for it are
%{\tt :exit\_stack\_minor -2} and
%{\tt :exit\_stack\_race -4} (for the former case),
%and
%{\tt :uses\_freed\_stack -4} (for the latter).
Whether each student's submitted P2 solved or suffered from this bug
was determined by me personally inspecting their code.

{\bf Mutex.}
The Landslide-friendly {\tt mutex\_test} checks for the possibility of two contending threads
accessing a mutex-protected critical section simultaneously.
It includes one thread repeating once the lock, unlock cycle
so as to check for unlock/lock races as well as lock/lock interactions,
and, as mentioned previously (\sect{\ref{sec:education-pebbles-tests}}),
checks for data race access pairs inside the mutex implementation itself.
As students have free rein to design their mutex internals,
this refers to the general class of mutex bugs
in which any number of things can go wrong depending on the implementation.
Whether each student's submitted P2 solved or suffered from this bug
was determined by checking their grade files for any mutex-related penalties (assessed by the TAs),
then me double-checking their implementations by hand to confirm.

Further investigating the one group who submitted a buggy mutex,
I found that while they had run {\tt mutex\_test} in Landslide
(and even found and fixed a separate deterministic bug already),
Landslide found no bug in what was presumably a correct implementation,
then they updated % oops
that implementation, introducing the bug, without testing it again thereafter.

@@

{\bf R/W-lock downgrade.}
In addition to the standard R/W-lock interface,
P2 requires students to implement {\tt rwlock\_downgrade()}:
called with the lock held in write mode,
the caller adopts the reader role instead,
allowing other waiting reader threads to proceed simultaneously,
all while allowing no waiting writers to access in between.
%
The Landslide-friendly test {\tt rwlock\_downgrade\_read\_test}
checks that readers are allowed simultaneous access after a downgrade with no possibility for deadlock.
Rather than one specific bug, this refers to any of several failures that can arise during a downgrade.
%, and
%the official rubric penalties related to this bug are
%{\tt :downgrade\_no\_signal -1},
%%{\tt :unsafe\_rwlock\_downgrade -2}, % don't think this is right; not relevant anyway (overlaps)
%{\tt :downgrade\_partial\_signal -1}, % not sure if this means smth that dgr would expose
%{\tt :downgrade\_\allowbreak{}lies -3}, and
%{\tt :downgrade\_deadlock -4}.
Whether each student's submitted P2 solved or suffered from these bugs
was determined by checking their grade files for any downgrade-related penalties
(which were assessed by the TAs as a result of their manual inspection, rather than mine).
